---
output:
  github_document
header-includes:
   - '<style>
        table img {
          background-color: transparent;
        }
      </style>'
---

# ggmol

ggmol is an R package designed for creating publication ready visualizations based on small molecules within the ggplot2 framework. The packages is built over ChemmineR and allows to simplify visualization by only relying on the molecules SMILES strings.

In the following code, the necessary libraries are loaded and the environment is set up for generating molecular visualizations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", dpi=300, out.width = 600)
library(ggmol)
library(ChemmineR)
library(ggforce) ## for annotating plots
library(umap)
library(MASS) ## to create a 2d map
set.seed(2)
```

A set of example molecules is defined using their SMILES representations and a `Chemmol` object is created, which is a container for handling chemical data in the ggmol package.

```{r}
example_set <- c(
Diazepam = "CN1c2ccc(Cl)cc2C(=NCC1=O)c1ccccc1",
Lorazepam = "OC1N=C(c2ccccc2Cl)c2cc(Cl)ccc2NC1=O",
Clonazepam = "[O-][N+](=O)c1ccc2NC(=O)CN=C(c3ccccc3Cl)c2c1",
Lisinopril = "NCCCC[C@H](N[C@@H](CCc1ccccc1)C(O)=O)C(=O)N1CCC[C@H]1C(O)=O",
Enalapril = "CCOC(=O)[C@H](CCc1ccccc1)N[C@@H](C)C(=O)N1CCC[C@H]1C(O)=O",
Ramipril = "CCOC(=O)[C@H](CCc1ccccc1)N[C@@H](C)C(=O)N1[C@H]2CCC[C@H]2C[C@H]1C(O)=O"
)

chemms <- Chemmol(smiles=new("SMIset", smilist=as.list(example_set)))
```

## Drawing structures

```{r}
ggplot(mapping=aes(0, 0, smiles=example_set[4], chemsize=2/3)) +
  geom_mol(show_atoms=T)+
  theme(aspect.ratio=1)+
  labs(title=names(example_set)[4], x=NULL, y=NULL)
```


In the following code, molecular glyphs are generated and plotted for each molecule in the example set. The `geom_mol()` function is used to render these glyphs within ggplot2, and `facet_wrap` arranges them to display each molecule separately.

```{r}
tab <- data.frame(
  name = names(example_set),
  smiles = example_set
)

ggplot(tab, aes(0, 0, smiles=smiles, label=name, chemsize=2/3)) +
  geom_mol()+
  theme_no_axes()+
  facet_wrap(~name)+
  theme(aspect.ratio=1)+
  labs(title="Faceted molecules")
```

## 2D embedding based on structural distance

A distance matrix for the molecules is computed using the ggmol `dist` function, and the Sammon mapping algorithm is used to generate a 2D embedding based on their structural similarities. K-means clustering is applied to group the molecules into clusters.

```{r}
moldists <- dist(chemms)
fprints_2d <- sammon(moldists)$points
clusters <- kmeans(fprints_2d, 2)$cluster

tab <- data.frame(
  x = fprints_2d[,1],
  y = fprints_2d[,2],
  name = names(example_set),
  smiles = example_set,
  Cluster= as.factor(clusters)
)
```


The following code shows a 2D embedding of the molecules, with each point representing a molecule and colored by its cluster. The `geom_mark_ellipse` function adds ellipses around clusters to highlight them.

```{r}
ggplot(tab, aes(x, y, smiles=smiles, colour=Cluster, chemsize=.1)) +
  geom_mol(show_atoms=F, resolution=700)+
  theme_bw()+
  labs(title="Molecules embedding with clusters")+
  lims(x=c(-100, 100), y=c(-35, 35))+
  geom_mark_ellipse(aes(colour=Cluster, smiles=NULL), expand = .07)+
  theme(aspect.ratio=1)
```

In the following code, labels are added to the molecules in the 2D embedding, making it easier to identify them. The `geom_label_repel` function helps to avoid overlapping labels.

```{r}
library(ggrepel)
ggplot(tab, aes(x, y, smiles=smiles, colour=Cluster, label=name, chemsize=.08)) +
  geom_mol(resolution=700, show_atoms=F)+
  labs(title="Molecules embedding with labels")+
  geom_label_repel(size=2.5, point.padding=17)+
  lims(x=c(-100, 100), y=c(-35, 30))+
  theme(aspect.ratio=1)
```

## Ranking molecules

The following code demonstrates how to rank molecules based on their distance scores. The bar plot shows the scores for each molecule, with the molecular glyphs displayed above the bars.

```{r}
scores <- as.matrix(moldists)[1,]

ggplot(tab[order(scores),], aes(x=1:nrow(tab), y=sort(scores)+1, smiles=smiles,
                label=name, chemsize=.1, fill=Cluster, colour = Cluster)) +
  geom_bar(stat="identity")+
  geom_mol(y=sort(scores)+13, show_atoms=F, resolution=700) +
  theme_classic()+
  labs(title="Molecules ranking", y="Scores", x="Molecules")+
  ylim(0, 160)+
  theme(aspect.ratio=1)+
  scale_x_continuous(breaks=1:nrow(tab), labels=tab$name)
```

## Table

In the following code, molecules are displayed side by side with their respective scores annotated. The `facet_wrap` function is used to arrange each molecule in a separate panel.

```{r}
# Create a data frame with molecule names and their corresponding SMILES strings
tab <- data.frame(
  name = names(example_set),
  smiles = example_set,
  row.names = NULL
)

# Generate molecular glyphs and store file paths
image_files <- generate_chem_glyphs(tab$smiles)


# Convert the image file paths to Markdown image links
#tab$image <- sprintf("![%s](%s)", tab$name, image_files)
tab$image <- sprintf('<img src="%s" alt="%s" width="100"/>', image_files, tab$name)


# Create a Markdown table
knitr::kable(tab, format = "markdown", col.names = c("Name", "SMILES", "Structure"))

```

